The Design and Implementation of the Synchronous Language \CEU

### 2016-03-30 paper.tex
...

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish with a summary of our design
    (\ref{sec.ceu.discussion}).

@@

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
events \code{B} and \code{C} are simultaneous, even though \code{A} and 
\code{B} actually happen much closer to one another.
In Section~\ref{sec.ceu.discussion}, after introducing internal events and 
first-class timers, we show an example that detects simultaneous button clicks.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about the program and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with intra-reaction communication, \CEU supports internal events which 
programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
Example \ax is safe because the trails access \code{x} atomically in different 
reactions;
Example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
the \C compiler that generates the final binary.
Therefore, access to \C is seamless and, more importantly, easily trackable.
%
Similarly to Esterel, which supports the \code{call} primitive for external 
code, calls are assumed to be instantaneous~\cite{esterel.primer}.
%
This way, programs should only resort to \C for asynchronous functionality, 
such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes (the compiler can restrict the available symbols in \C).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{\ax Definition and use of external symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{\bx Annotations for the external symbols in \ax.}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The unsafe program in \ax only compiles with the annotations in \bx.
\label{lst.c}
}
\end{figure}

We extend the temporal analysis of \CEU for calls and accesses to external 
symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block with standard declarations in \C (ln.  
1--6).
%
The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
during the boot reaction (ln. 8,10):
the first trail calls the symbol \code{\_f},
while the second calls \code{\_g} and \code{\_id}, and also accesses 
\code{\_NUM}.
%
Since \CEU does not inspect any code defined in \C, it complains about 
suspicious concurrent accesses between \code{\_f} and all symbols in the second 
trail.

The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
compiler about the semantics of the \C symbols in use in \ax:
%
\begin{itemize}
\item \code{\_NUM} is a constant symbol, meaning that it is safe to use it 
      concurrently with any other symbol in the program.
\item \code{\_id} is a pure function, also meaning that it is safe to call it
      concurrently with any other symbol in the program.
\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
      conflict and they can be safely called concurrently.
\end{itemize}

With the annotations in \bx, the program in \ax compiles successfully.

Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish how 
synchronous languages can abort awaiting lines of execution without tweaking 
them with synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
%
Still, aborting lines of execution that deal with external resources is 
challenging because they may end up in an inconsistent state.
%
For this reason, Esterel and \CEU provide a \code{finalize} construct to 
unconditionally execute a series of statements even if the enclosing block is 
aborted and does not terminate normally.

@@

\subsection{ Discussion }
\label{sec.ceu.discussion}

TODO: queue-based/unique in the semantics of the language and not in the 
library

permeates all aspects

\CEU aims to offer a simpler semantics than Esterel while keeping the power of 
synchronous languages, with a tractable concurrency model, reactive and 
deterministic execution, and safe support for orthogonal abortion of lines of 
execution.

With the event-based semantics, \CEU does not impose a notion of simultaneity 
for events, requiring each application to define it only when necessary.
%
Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
\code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Note also that the application in \CEU could define an external input 
\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
10) to recover the abstract notion of simultaneity.

@@

internal events
    - flexible, coroutines, resumable exceptions
    - yet simple, programmers understand stacks
    - most of the time doesn't matter

@@

@@

@@

>>>>

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
%We finish with a summary of our design
    %(\ref{sec.ceu.discussion}).

@@

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
events \code{B} and \code{C} are simultaneous, even though \code{A} and 
\code{B} actually happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example that detects simultaneous button clicks.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about the program and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with intra-reaction communication, \CEU supports internal events which 
programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), broadcasts \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item No other trails awake to \code{b}.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are safe}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are unsafe}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
Example \ax is safe because the trails access \code{x} atomically in different 
reactions;
Example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
the \C compiler that generates the final binary.
Therefore, access to \C is seamless and, more importantly, easily trackable.
%
Similarly to Esterel, which supports the \code{call} primitive for external 
code, calls are assumed to be instantaneous~\cite{esterel.primer}.
%
This way, programs should only resort to \C for asynchronous functionality%
\footnote{
In \CEU, it is possible to restrict the available \C symbols as a compile-time 
option.
}, such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{\ax Definitions and uses of symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{\bx Annotations for the symbols in \ax}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The unsafe program in \ax only compiles with the annotations in \bx.
\label{lst.c}
}
\end{figure}

The temporal analysis of Section~\ref{sec.ceu.analysis} also considers calls 
and accesses to external symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block with standard declarations in \C (ln.  
1--6).
%
The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
during the boot reaction (ln. 8,10):
the first trail calls symbol \code{\_f},
while the second calls \code{\_g} and \code{\_id}, and also reads \code{\_NUM}.
%
Since \CEU does not inspect any code in \C, it complains about suspicious 
concurrent accesses between \code{\_f} and all symbols in the second trail.

The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
compiler about the semantics of the \C symbols in use in \ax.
With the annotations in \bx, the program in \ax compiles successfully:
%
\begin{itemize}
\item \code{\_NUM} is a constant symbol, meaning that it is safe to use it 
      concurrently with any other symbol in the program.
\item \code{\_id} is a pure function, also meaning that it is safe to call it
      concurrently with any other symbol in the program.
\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
      conflict and they can be safely called concurrently.
\end{itemize}

Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish 
synchronous from asynchronous languages with respect to orthogonal abortion of
lines of execution (i.e., without tweaking them with synchronization 
primitives)~\cite{esterel.preemption}.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{sync_async.threadsstop}.
%
Still, aborting lines of execution that deal with external resources is 
challenging because they may end up in an inconsistent state.
%
For this reason, Esterel and \CEU provide a \code{finalize} construct to 
unconditionally execute a series of statements even if the enclosing block is 
aborted and does not terminate normally.

@@

In Section~\ref{sec.ceu.exts}, we argue that events occurrences are 
infinitesimal and can never be absolutely simultaneous.
However, the ``sensation of simultaneity'' is not infinitesimal, but increases 
with the inaccuracy of the observer (e.g., a human being).
Therefore, simultaneity should be defined with respect to the observer, which 
varies from case to case.

First-class timers simplify the implementation of application-defined 
simultaneity.
%
Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
\code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Note also that the application in \CEU could define an external input 
\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
10) to recover the abstract notion of simultaneity.

@@

@@

The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.

@@

\begin{comment}
%\subsection{ Discussion }
\label{sec.ceu.discussion}

TODO: related
- simplicity
    - intra-reaction det
    - internal events?
        - more fine grained

% SIMPLE

\CEU aims to offer a simpler semantics than Esterel without giving up its 
tractable synchronous concurrency model:

% unique external events

% internal events

Internal events

% lexical order

    - flexible, coroutines, resumable exceptions
    - yet simple, programmers understand stacks
    - most of the time doesn't matter

Furthermore, our broad safety analysis, encompassing all proposed concurrency 
mechanisms, ensures that the high degree of concurrency in WSNs does not pose 
safety threats to applications.
%
%Even with the restrictions that were needed to enable this analysis, \CEU was 
%shown to be expressive enough to imply in a decrease in code complexity for a 
%range of typical WSN applications.
%
%By restricting the expressiveness of the language, we could enable a broad 
%safety analysis in programs at compile time.
%The analysis encompasses all proposed concurrency mechanisms, such as parallel 
%compositions, first-class timers, and communication via events.
%

% SAFE

Esterel XXX
thanks to the uniqueness of events, pointer tracking, and first-class timers, 
\CEU also provides additional safety protections:
%
No concurrency in accesses to shared variables and system calls sharing a 
resource
Finalization for blocks going out of scope
Auto-adjustment for timers in sequence and synchronization for timers in parallel

These properties are desirable in any application and are guaranteed as 
preconditions in \CEU by design.
Ensuring or even extracting these properties from less restricted languages 
requires significant manual analysis.

% LIMITATIONS

Even though the achieved expressiveness and overhead of \CEU meet the 
requirements of WSNs, its design imposes two inherent limitations:
the lack of dynamic loading which would forbid the static analysis,
and the lack of hard real-time guarantees.

Regarding the first limitation, dynamic features are already discouraged due to 
resource constraints.
For instance, even object-oriented languages targeting WSNs forbid dynamic 
allocation~\cite{wsn.flowtalk,wsn.virgil}.
%
Given that we focus on system-level development which does not require rich 
dynamic functionality, we leave for a future work an in-depth discussion about 
this issue.
Note that queues, stacks, and other simple dynamic data structures for handling 
message packets can be made available to \CEU through its safe integration with 
$C$.

To deal with the second limitation, which can be critical in the presence of 
lengthy computations, we can consider the following approaches:
(1) manually placing \code{pause} statements in unbounded loops;
(2) integrating \CEU with a preemptive system.
%
The first option requires the lengthy operations to be rewritten in \CEU using
\code{pause} statements so that other trails can be interleaved with them.
This option is the one recommended in many related work that provide a similar 
cooperative primitive (e.g. \code{pause}~\cite{esterel.primer}, 
\code{PT\_YIELD}~\cite{wsn.protothreads}, \code{yield}~\cite{wsn.sol}, 
\code{post}~\cite{wsn.nesc}).
%
Considering the second option, \CEU and preemptive threads are not mutually 
exclusive.
% and the second option is appealing to be explored in a future work.
For instance, TOSThreads~\cite{wsn.tosthreads} proposes a message-based 
integration with \emph{nesC} that is safe and matches the semantics of \CEU 
external events.
\end{comment}

@@

% CONTRIBS
    - queue in the semantics
    - permeates all aspects
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library
